---
title: "EpiData-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiData-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This package is an alternative missing data imputation pipeline for bisulfite sequencing. 

Accurate imputation of missing methylation values is a crucial step in bisulfite sequencing (BS-seq) analysis, as incomplete data can bias downstream association or differential methylation studies.
EpiData provides a modular and reproducible framework for preprocessing and imputation of BS-seq data.


The main objectives of EpiData are:

- to ensure data integrity and quality before statistical modeling

- to provide parallelized and scalable computation for large methylome datasets

- to allow flexible integration with downstream pipelines (e.g., differential methylation, EWAS, or multi-omics integration)

### Loading the package:

```{r setup}
library(EpiData)
```



This vignette demonstrated a complete pipeline for:

- Cleaning coverage and removing outliers

- Computing methylation values

- Calculating CpG-level statistics

- Imputing missing values and combining results


We will go through each function step-by-step, from filtering and outlier removal to computing CpG statistics and imputing missing data. The workflow ensures high-quality methylation matrices ready for downstream analysis such as differential methylation or association studies.

### Data import and description

```{r}
data("bs_list")
```


The object bs_list contains example methylation data. It includes three assays: the total coverage count matrix, the methylated counts matrix, the unmethylated counts matrix for 150 CpG (150 rows) and 10 individuals. 


Otherwise own data can be use, derived from object class such as **SummarizedExperiment** object.

## Workflow

This vignette guides you through the complete imputation workflow, demonstrating how to:

- Clean coverage and remove outliers

- Compute methylation proportions

- Summarize CpG-level statistics

- Impute missing values using correlation-based methods

- Combine results into ready-to-analyze methylation matrices

The resulting datasets are high-quality, imputed methylation matrices suitable for downstream analyses such as differential methylation testing, clustering, or predictive modeling.

### Step 1: Cleaning the coverage matrix

We start by removing CpGs or individuals with too many missing values using **cleanCovMat()**.
Missing or low-coverage CpGs can distort methylation estimates.


```{r}
clean.coverage <- cleanCovMat(
input.obj = bs_list,
max_na_cpg = 0.5, # max 50% missing per CpG
max_na_ind = 0.2, # max 20% missing per individual
cpg_removal_threshold = 10 # minimum number of valid CpGs required
)
```

This step returns a cleaned coverage object where low-quality CpGs or samples are excluded.
The function automatically retains metadata compatibility with subsequent steps.


### Step 2: Outlier detection and removal

Extreme methylation or coverage values may arise from technical noise or sequencing artifacts.
**cleanOutliers()** detects and optionally removes these points based on customizable thresholds.

```{r}
clean.out <- cleanOutliers(
filtered.obj = clean.coverage,
outlier_threshold = 0,
remove_outliers = TRUE
)
```

This step ensures that methylation values reflect biological rather than technical variation.

### Step 3: Compute methylation values

Methylation proportions are calculated as the ratio of methylated counts to total coverage.
Convert the cleaned object into methylation proportion matrices with **methValues()**.

```{r}
mat.values <- methValues(clean.coverage)
```

### Step 4: Compute CpG-level statistics

CpG-wise summary statistics (mean, variance, missingness) are computed in parallel using **statsCpG_parallel()**.
This step provides an overview of methylation distribution and reliability across the dataset.


```{r}
stat.result <- statsCpG_parallel(mat.values)
```

The output is a list of chromosome-specific results, allowing scalable analysis of genome-wide data.


### Step 5: Impute missing data

Missing methylation values are imputed using correlation-based imputation via **imputeCorr_parallel()**.
This approach uses CpG-level correlations and local methylation patterns to predict missing entries, preserving biological structure.

```{r}
imputed <- imputeCorr_parallel(stat.result)
```

Each chromosome’s imputed object retains both raw and imputed methylation matrices, enabling traceability.


### Step 6: Combine imputed data frames

Each chromosome result includes an imputed data frame.
We combine all imputed data frames into a single object using the custom helper function below.

```{r}
combine_imputed <- function(chr_list) {
  imputed_list <- lapply(chr_list, function(x) x[["imputed"]]) 
  do.call(rbind, imputed_list)                                  
}

# Example use:
combined_imputed_df <- combine_imputed(imputed)
```

This unified imputed dataset is now ready for downstream statistical analysis and visualization.






