---
title: "Missing data imputation pipeline for Bisulphite Sequencing analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data imputation pipeline for BS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 

`EpiData`  is an R package designed to enhance the reliability and reproducibility of DNA methylation analysis. It provides tools to normalize M-value distributions per CpG site, improving data comparability across samples, and introduces a novel approach for imputing missing values in bisulfite sequencing experiments.


## Installation and required packages

```{r, warning=FALSE}
library(EpiData)
library(SummarizedExperiment)
```

# Get started with the analysis

## Loading data 
```{r}
data("meth_data")
```

The data contains a list of assays from experiment obtained by bisulfite sequencing 43 individuals. 

As an example data derives from chromosome 22 (the same pipeline is applied to the whole chromosome matrix if it is provided): 


The dataset contains the LargeSummarizedExperiment object with the following list of assays from experiment obtained by bisulfite sequencing 43 individuals. As an example data derives from chromosome 22: 

- coverage counts matrix (Coverage_matrix)

- methylated counts matrix (Met_matrix)

- unmenthylated counts matrix (Unmet_matrix)

In this objects exact locations and CpGs' names are present (labeled as chr:bp). 

Assume that from the experiment we endend up with a dataset not only contanining chromosome 22, the optimal situation would be split dataset per each chromosome and perform a parallelized analysis. 



## Analysis part 1: data set and cleaning the whole matrix of the CpGs

```{r}
input.list<- assays(meth_data)
```

**NOTE** that if you start from three raw matrices (coverage matrix, methylated and unmethylated counts matrices) -not arranged as GRanged object- you can put them in a list of three dataframe as shown: 

```{r, warning=FALSE}
input.list<- SummarizedExperiment::assays(meth_data)
head(input.list[[1]]) # coverage matrix
head(input.list[[2]]) # methylated counts matrix
head(input.list[[3]]) # unmethylated counts matrix
```

And then run directly the passage of the cleaning step 1 (Pipeline Part 1).


**Cleaning step 1: coverage**

In this step, we clean the input data by applying coverage filters. Specifically, we remove CpGs and individuals with excessive missing values and exclude CpGs with very low coverage (fewer than 10 reads by default).

Coverage filtering ensures that downstream analyses are based on reliable and comparable data. CpGs with low coverage or individuals with too many missing values can introduce noise and bias, potentially leading to spurious results. By applying these thresholds, we retain high-quality CpG sites and individuals, improving the robustness of subsequent statistical analyses.

```{r}
clean.coverage <- cleanCovMat(input.obj = input.list, max_na_cpg = 0.5, max_na_ind = 0.2,  cpg_removal_threshold = 10)
```

**Cleaning step 2: outliers**

```{r}
clean.out<- cleanOutliers(filtered.obj=clean.coverage, outlier_threshold=0, remove_outliers = T)
```

## Analysis part 2: methylation values

Here are calculated the values that are used for testing methylation data, i.e., the beta values and the M values from the matrices of counts data.

```{r}
mat.values <- methValues(clean.coverage)
```

## Analysis part 3: imputation

### Calculating statistics per CpG and imputing data

```{r}
stat.result <- missingStats(list(mat.values))
```

### Imputing data

Our method imputes missing values in methylation datasets by computing the correlation structure between CpG sites. It first computes a non-negative definite covariance matrix via eigen-decomposition, then simulates new values from the same distribution to fill in missing entries. The approach is parallelized for efficiency and preserves the natural correlation patterns in the data, improving accuracy over standard imputation methods.


The method that we introduce, firstly fills missing values with row means. Secondly, it calculates a correlation matrix, it constructs a covariance matrix, ensuring it is positive semi-definite via eigen decomposition and finally simulates new values from a multivariate normal distribution to impute missing entries.

```{r}
imputed <- batchImputeCorr(stat.result)
imputed_data <- (imputed[[1]][[1]])

imputed_data[1:4, 1:10]

```

## Imputation comparison pipeline  

Comparison of this proposed method to other existing methods

This pipeline compares our new imputation method to existing approaches. 

Steps: 

Here are compared our new imputation method to existing approaches.  

- **Our method**: computes a non-negative definite covariance matrix via eigen-decomposition and simulates new values from the same distribution, preserving CpG correlation patterns (parallelized).  

- **Existing methods**: alternative imputation strategies included for benchmarking.  

- **Evaluation**: performance is assessed with RMSE and MAE after introducing artificial missingness.  


The function imputeCorr() implements a correlation-based imputation method for DNA methylation matrices, designed to recover missing CpG-level values from bisulfite sequencing or array-based data.
It extract the correlation structure between CpG sites to generate realistic imputed values consistent with the underlying methylation patterns.

```{r}
data("meth_data")
input.list <- assays(meth_data)
clean.coverage <- cleanCovMat(input.obj=input.list, max_na_cpg = 0.5,
                                  max_na_ind = 0.2,  cpg_removal_threshold = 10)

mat.values <- methValues(clean.coverage)
mat.values <- lapply(mat.values, na.omit)

imp.M <- repNA(cleaned.obj=mat.values,
                 missing_prop= 0.2,
                 varselect=5,
                 n.iter=4,
                 sel_method=c(1:3, 9:10),
                 trees= 50, nb= 10, ncomp= 2,
                 matrix="M")

measure_imp_m <- measureAccuracy(imp.M)
measure_imp_m$Boxplot.rmse
measure_imp_m$Boxplot.mae
```

### Rationale fot the proposed method: 

Why using correlation-based imputation for DNA methylation data.

- DNA methylation patterns are highly correlated within genomic regions, such as CpG islands, shores, enhancers, and gene bodies.
Neighboring CpG sites often share similar methylation dynamics because they are regulated by the same chromatin context, DNA-binding proteins, or local transcriptional activity.

- When missing values occur—due to sequencing depth variability, bisulfite conversion inefficiency, or array detection limits—naively replacing them with row means or most frequent values can distort this local structure.
Such substitutions:

- Ignore spatial and regulatory dependencies among CpGs.

Reduce biological variance and weaken downstream analyses (e.g., PCA, differential methylation testing).

- Introduce artificial clusters that do not reflect true methylation variability.

- In contrast, correlation-based imputation uses the empirical co-methylation structure across CpGs to estimate missing entries in a biologically informed way.

- By simulating values from a multivariate normal model constrained by the observed covariance, imputeCorr() preserves both:

- Global methylation trends across individuals, and

- Local co-regulation between CpGs within regions.

- This results in more realistic methylation matrices for downstream modeling, differential testing, and network inference.
