---
title: "EpiData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  ## edited options below
  fig.width = 8,
  fig.asp = 0.8,
  out.width = "100%",
  dpi = 300,
  warning = FALSE, 
  message = FALSE, 
  echo = TRUE) 
```


## Loading the package and useful libraries
```{r, warning=FALSE}
library(EpiData)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(purrr)
library(dplyr)
library(SummarizedExperiment)
library(GenomicRanges)
library(GenomicFeatures)
library(MASS)
library(stats)
```



## Loading data 
```{r}
data("rangedObject")
```


The dataset contains the LargeSummarizedExperiment object with the following list of assays from experiment obtained by bisulfite sequencing 43 individuals. As an example data derives from chromosome 22: 

- coverage counts' matrix (Coverage_matrix)

- methylated counts' matrix (Met_matrix)

- unmenthylated counts' matrix (Unmet_matrix)

In this objects exact locations and CpGs' names are present (labeled as chr:bp). 

Assume that from the experiment we endend up with a dataset not only contanining chromosome 22, the optimal situation would be split dataset per each chromosome and perform a parallelized analysis. 


Note that if data is not in this format and sequencing gave the three matrices, this operation can be done with: 
```{r}
n_sites <- 10
n_indiv <- 6

# Generate random data for methylated counts
set.seed(123)  # For reproducibility
methylated <- as.data.frame(matrix(sample(0:100, n_sites * n_indiv, replace = TRUE), nrow = n_sites, ncol = n_indiv))

# Generate random data for unmethylated counts
unmethylated <- as.data.frame(matrix(sample(0:100, n_sites * n_indiv, replace = TRUE), nrow = n_sites, ncol = n_indiv))

# Total coverage is the sum of methylated and unmethylated counts
coverage <- as.data.frame(methylated + unmethylated)

input.list <- list(Coverage_matrix=coverage, Met_matrix=methylated, Unmet_matrix=unmethylated)
```


Without the ranged object the final input list can be obtained with `GRconvresion` function

```{r}
gr.object <- GRconversion(input.list)

input.list <-  assays(gr.object)
```


Otherwise the ranged object can be used:
```{r message=FALSE, warning = FALSE}

# with the ranged object
input.list <- assays(rangedObject)

subset_list <- lapply(input.list, function(df) df[1:100, 1:10])
```


# Pipeline

## First part: filtering

### Filtering 1
```{r message=FALSE, warning = FALSE}
clean.coverage <- cleanCovMat(input.obj = subset_list, max_na_cpg = 0.5, max_na_ind = 0.2,  cpg_removal_threshold = 10)
```


Function `cleanCovMat` filters coverage matrix, consideriign the following assumptions as default: 

1) if the coverage counts are zero, then NAs are put, because coverage is the sum given by unmethylated counts and methylated counts, hence it should not be equal to zero;


2)	CpGs (rows) with missing values greater than 50% are removed;

3)	individuals (columns) with a missing percentage greater than 20% are removed.

### Filtering 2: beta and M values

```{r message=FALSE, warning = FALSE}
clean.out <- cleanOutliers(filtered.obj=clean.coverage, outlier_threshold=0, remove_outliers = TRUE)
```



The function returns filtered counts' matrices for excessive NAs in methylated and unmenhylated counts' matrices. 

- `cleanOutliers` filters per rows (CpGs) and columns (individuals). First the function removes entries which may still have undesirable high percentages of NAs in methylated or unmethylated matrices, not yet cleaned from cleaning step 1. Secondly, the function clean outliers, since they can significantly impact the results of statistical analyses. 



```{r}
values <- MethilationValues(clean.out)
```

- Moreover this function computes the matrices of beta values and M values. 

However, issues may arise when converting outliers to NAs, especially in studies focusing on specific areas, patient categories, or low-quality sequencing results after quality control. On the other hand, coverage of sequencing should be considered when removing outliers from sequencing experiments to ensure the reliability of the analysis. Outliers, which are data points that appear not to belong among the other data present, can significantly impact statistical analyses. While it may seem simple to remove outliers, it is crucial to handle them appropriately due to their potential impact on results and assumptions. Coverage must be considered when approaching removing counts. 


After cleaning the data, the pipeline follows with studying each chromosome seprately. The following function helps in separating the information of BS experiment per each chromosome. 

### Computing statistics and imputation

The input should be the assay method of the SummarizedExperiment object, if present. Otherwise a list with the three matrices described before.

The result is a list where we have the total coverage matrix, methylated maytrix and unmethylated matrix with all chromoeomse, but also a list called *final* which contains all these three matrices already divided per each chromosome. 


## Second part: imputation
After cleaning steps, CpGs' matrices are divided per each chromosome with the function

```{r warning=FALSE, message=FALSE}
library(BiocParallel)
library(matrixcalc)
```


### Calculating statistics per CpG 
```{r warning=FALSE}
#clean.out2<- list(values, values)
#stats<- missingStats(clean.out2)
stats <- statsCpG(values)
```


### Imputing data based on correlation
```{r  warning=FALSE}
#simu<- eigenImpute(stats)
simu <- imputeCorr(stats)
```

Here is shown the example of `imputeCorr` function (used by ParallSimu) through each one of the chromosomes. Note that the input could be *pairwise* as well as *meanCpG*.

When computing correlation matrix in presence of missing data, the correlation pattern behind the data may result affected by NAs presence. Since out aim is to impute the M values matrix, the  of CpGs' correlation can be reconstructed via computing the correlation matrix once imputed the mean vM value per each CpG, not to change the similarity between CpGs or via pairwise correlation. 


# Comparison of imputation methods
```{r, echo=F}
data.full <- lapply(values, na.omit)
imp.M <- repNA(cleaned.obj=data.full,
                missing_prop= 0.2,
                varselect=5,
                n.iter= 5,
                sel_method=c(1:10) ,
                trees= 50, nb= 10, 
                ncomp= 2,
                matrix="M")

```


## Measuring accuracy
```{r}
measures<- measureAccuracy(imp.M)
```


**Measuring RMSE**
```{r}
measures$Boxplot.rmse
```



**Measuring MAE**
```{r}
measures$Boxplot.mae
```

