---
title: "EpiData-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiData-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```


## Introduction

This package is an alternative missing data imputation pipeline for bisulfite sequencing. 

Accurate imputation of missing methylation values is a crucial step in bisulfite sequencing (BS-seq) analysis, as incomplete data can bias downstream association or differential methylation studies.
EpiData provides a modular and reproducible framework for preprocessing and imputation of BS-seq data.


The main objectives of EpiData are:

- to ensure data integrity and quality before statistical modeling

- to provide parallelized and scalable computation for large methylome datasets

- to allow flexible integration with downstream pipelines (e.g., differential methylation, EWAS, or multi-omics integration)

# Description of the package

Welcome to the `EpiData` project, which is a package for enhancing reliability in DNA methylation analysis (contributing to normalize M value disstribution per CpG) and propose a novel approach for imputing missing data from bisulfite sequencing experiments. 
Accurate imputation of missing methylation values is a crucial step in bisulfite sequencing (BS-seq) analysis, as incomplete data can bias downstream association or differential methylation studies.
EpiData provides a modular and reproducible framework for preprocessing and imputation of BS-seq data.

The main objectives of EpiData are:

- to ensure data integrity and quality before statistical modeling

- to provide parallelized and scalable computation for large methylome datasets

- to allow flexible integration with downstream pipelines (e.g., differential methylation, EWAS, or multi-omics integration)


## Installing the package

Please install devtools and Bioconductor if you haven't them yet.

# Introduction

This is a short vignette to give an example about how to read bisulfite sequencing data into `EpiData` package. 

Let's load the package first. 

```{r library}
library(EpiData)
```

# Preparing data 

## Reading data

The `EpiData` uses [`BSseq`](https://www.bioconductor.org/packages/release/bioc/vignettes/bsseq/inst/doc/bsseq.html#3_Using_objects_of_class_BSseq) object as the primary data input. Please refer to [this guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/bsseq/inst/doc/bsseq.html) for details about how to create and manipulate `BSseq` objects. Below is an example from our package:

```{r read_data, warning=F}

# Data
data_dir <- system.file("extdata", package = "EpiData")
cov_files <- list.files(data_dir, pattern = "\\.cov\\.gz$", full.names = TRUE)

# Metadata (not necessary, but good to have)
data_dir <- system.file("extdata", package = "EpiData")
metadata_file <- list.files(data_dir, pattern = "\\.tsv\\.gz$", full.names = TRUE)

metadata <- read.table(metadata_file, header = TRUE)
rownames(metadata) <- metadata[, 2]

# BSseq object
bs <- bsseq::read.bismark(files = cov_files, 
                          colData = metadata, 
                          strandCollapse = TRUE,
                          rmZeroCov = TRUE)
bs
```

## Filtering missing data

A high-quality set of CpG sites is essential before carrying out downstream imputation or modeling. To achieve this, we first retain CpGs that meet a minimum coverage threshold (`min_cov`) in at least a specified proportion of samples (`row_min`). After filtering the CpGs, we then retain only those samples that satisfy the same coverage requirement in at least a specified proportion of the remaining CpGs (`col_min`). This two-step filtering strategy ensures that both rows (CpGs) and columns (samples) contain sufficient, high-confidence coverage information, providing a robust foundation for subsequent imputation and analysis.

```{r filter_data}
# Filter low-coverage CpGs and samples
bs_filt <- filterCov(bs = bs,
                     min_cov = 10,
                     row_min = 0.5, 
                     col_min = 0.5)
bs_filt
```

## Preparing `BSseq` object

Before we proceed with the imputation step, we need to make sure that the CpG sites that has zero coverage is actually a missing value and therefore treated as `NA`.

```{r process_data}
bs_filt <- processBSseq(bs_filt)
```



## Preparing the required matrices

Various matrices are required for downstream imputation process. It includes:

| Matrix / Object | Description                                                                       |
|:----------------|:----------------------------------------------------------------------------------|
| `coverage`      | Total CpG coverage                                                                |
| `meth`          | Number of methylated reads                                                        |
| `unmeth`        | Number of unmethylated reads                                                      |
| `beta`          | Methylation percentages                                                           |
| `m`             | M-values per CpG per sample, calculated as `log2(meth/unmeth)`                    |
| `gr`            | `GRanges` object containing genomic coordinates (chromosome, start, end, strand)  |

```{r calc_matrices}
bs_mat <- extractMethData(bs_filt)
lapply(bs_mat, head, 5)
```


```{r}

m_mat <- bs_mat$m
gr<- bs_mat$gr

positions <- data.frame(
  CpG = rownames(m_mat),      
  chr = GenomicRanges::seqnames(gr),
  pos = GenomicRanges::start(gr),
  stringsAsFactors = FALSE)


res <- positionClusterImpute(
  m_values = m_mat,
  positions = positions,
  dist_threshold = 1000,
  impute_method = "mean"   # or "knn"
)

```




```{r}
report_qc<-methylationQC(res,
                          beta_values = NULL,
                          positions = NULL,
                          sample_metadata = NULL,
                          max_na_per_probe = 0.2,
                          max_na_per_sample = 0.1,
                          m_range = c(-10, 10),
                          detect_outliers = TRUE,
                          outlier_sd = 3,
                          verbose = TRUE) 
  
```



