---
title: "EpiData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(EpiData)
```



## Required packages
```{r, warning=FALSE, message=FALSE}
library(devtools)
library(magrittr)
library(purrr)
library(dplyr)
library(psych)
library(kableExtra)
library(GenomicRanges)
library(SummarizedExperiment)
library(radiant.data)
```



## Loading data 
```{r}
data("matrices")
```


The dataset contains the LargeSummarizedExperiment object with the following list of assays from experiment obtained by bisulfite sequencing 43 individuals. As an example data derives from chromosome 22: 

- coverage counts' matrix (Coverage_matrix)

- methylated counts' matrix (Met_matrix)

- unmenthylated counts' matrix (Unmet_matrix)

In this objects exact locations and CpGs' names are present (labeled as chr:bp). 

Assume that from the experiment we endend up with a dataset not only contanining chromosome 22, the optimal situation would be split dataset per each chromosome and perform a parallelized analysis. 



# Pipeline

## Cleaning step 1

```{r message=FALSE, warning = FALSE}
input.list<- assays(dati)

clean.coverage <- cleanCovMat(input.obj = input.list, max_na_cpg = 0.5, max_na_ind = 0.2,  cpg_removal_threshold = 10)
```


Function `cleanCovMat` filters coverage matrix, consideriign the following assumptions as default: 

1) if the coverage counts are zero, then NAs are put, because coverage is the sum given by unmethylated counts and methylated counts, hence it should not be equal to zero;


2)	CpGs (rows) with missing values greater than 50% are removed;

3)	individuals (columns) with a missing percentage greater than 20% are removed.


## Cleaning step 2

```{r message=FALSE, warning = FALSE}
clean.out<- cleanOutliers(filtered.obj=clean.coverage, outlier_threshold=0, remove_outliers = T)

```

The function returns filtered counts' matrices for excessive NAs in methylated and unmenhylated counts' matrices. 

`cleanOutliers` filters per rows (CpGs) and columns (individuals). First the function removes entries which may still have undesirable high percentages of NAs in methylated or unmethylated matrices, not yet cleaned from cleaning step 1. Secondly, the function clean outliers, since they can significantly impact the results of statistical analyses. 


However, issues may arise when converting outliers to NAs, especially in studies focusing on specific areas, patient categories, or low-quality sequencing results after quality control. On the other hand, coverage of sequencing should be considered when removing outliers from sequencing experiments to ensure the reliability of the analysis. Outliers, which are data points that appear not to belong among the other data present, can significantly impact statistical analyses. While it may seem simple to remove outliers, it is crucial to handle them appropriately due to their potential impact on results and assumptions. Coverage must be considered when approaching removing counts. 


After cleaning the data, the pipeline follows with studying each chromosome seprately.the following function helps in separating the information of BS experiment per each chromosome. 

### Splitting dataset per chromosomes

The input should be the assay method of the SummarizedExperiment object, if present. Otherwise a list with the three matrices described before.
```{r}
splitted<- split5MatXChrom(clean.out)
```

The result is a list where we have the total coverage matrix, methylated maytrix and unmethylated matrix with all chromoeomse, but also a list called *final* which contains all these three matrices already divided per each chromosome. 


## Parallelization
After cleaning steps, CpGs' matrices are divided per each chromosome with the function

```{r warning=FALSE, message=FALSE}
library(BiocParallel)
library(matrixcalc)
```


### Calculating statistics per CpG 
```{r warning=FALSE}
input.stats<- splitted$final

# Statistics
stats2<- ParallStats(input.stats)
```


Here is shown the example of `statsCpG` function (used by ParallStats) through each one of the chromosomes. 

```{r message=FALSE, warning = FALSE}
final2<- splitted$final

stats<- statsCpG(cleaned.obj= final2[[1]], varselect = 5, plot = TRUE)
stats$Barplot
```



### Imputing data based on correlation
```{r  warning=FALSE}
simu2<- ParallSimu(stats2)
```

Here is shown the example of `imputeCorr` function (used by ParallSimu) through each one of the chromosomes. Note that the input could be *pairwise* as well as *meanCpG*.


Imputation based on correlation structure 
```{r message=FALSE, warning = FALSE}
obj.imp.pairw <- imputeCorr(cleaned.obj=stats,
                    matrix="M",
                    varselect = 5,
                    correlation.type="pairwise")


obj.imp.mean <- imputeCorr(cleaned.obj=stats,
                    matrix="M",
                    varselect = 5,
                    correlation.type="meanCpG")


```


The default missing values' investigation is performed for M values. Note that percentage of missing values in M values are higher than counts' missing value percentages.


# Comparison of imputation methods

```{r, message=FALSE, warning=FALSE}
stats3<- lapply(clean.out, na.omit)

imp.M <- repNA(list.cleaned=stats3,
                       missing_prop= 0.2, varselect=5,
                       n.iter= 2, sel_method=c(1:10),
                       trees= 50, nb= 10, ncomp= 2, matrix="M")

```


```{r, warning=FALSE, message=FALSE}
measure_imp_m <- measureAccuracy(imp.M)
measure_imp_m$Boxplot.rmse
measure_imp_m$Boxplot.mae
```


