---
title: "Reading data into EpiData"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Read Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a short vignette to give an example about how to read bisulfite sequencing data into `EpiData` package. 

Let's load the package first. 

```{r library}
library(EpiData)
```

# Reading data

The `EpiData` uses [`BSseq`](https://www.bioconductor.org/packages/release/bioc/vignettes/bsseq/inst/doc/bsseq.html#3_Using_objects_of_class_BSseq) object as the primary data input. Please refer to [this guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/bsseq/inst/doc/bsseq.html) for details about how to create and manipulate `BSseq` objects. Below is an example from our package:

```{r read_data}
# Data
data_dir <- system.file("extdata", package = "EpiData")
cov_files <- list.files(data_dir, pattern = "\\.cov\\.gz$", full.names = TRUE)

# Metadata (not necessary, but good to have)
data_dir <- system.file("extdata", package = "EpiData")
metadata_file <- list.files(data_dir, pattern = "\\.tsv\\.gz$", full.names = TRUE)
metadata <- read.table(metadata_file, header = TRUE)
rownames(metadata) <- metadata[, 2]

# BSseq object
bs <- bsseq::read.bismark(files = cov_files, 
                          colData = metadata, 
                          strandCollapse = TRUE,
                          rmZeroCov = TRUE)
bs
```

# Filtering missing data

A high-quality set of CpG sites is essential before carrying out downstream imputation or modeling. To achieve this, we first retain CpGs that meet a minimum coverage threshold (`min_cov`) in at least a specified proportion of samples (`row_min`). After filtering the CpGs, we then retain only those samples that satisfy the same coverage requirement in at least a specified proportion of the remaining CpGs (`col_min`). This two-step filtering strategy ensures that both rows (CpGs) and columns (samples) contain sufficient, high-confidence coverage information, providing a robust foundation for subsequent imputation and analysis.

```{r filter_data}
# Filter low-coverage CpGs and samples
bs_filt <- filterCov(bs = bs,
                     min_cov = 10,
                     row_min = 0.5, 
                     col_min = 0.5)
```

